---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r not-cran-check, include = FALSE}
# Skip code evaluation on CRAN (external API calls will fail)
not_cran_check <- identical(Sys.getenv("NOT_CRAN"), "true") &&
  !nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
run_vignette <- isTRUE(getOption("pkgdown.in_progress")) || not_cran_check
knitr::opts_chunk$set(eval = run_vignette, purl = run_vignette)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,        # standardized width that fits vignette text column
  fig.height = 5,       # default height, override in individual chunks as needed
  fig.dpi = 150,        # higher resolution for crisp rendering
  out.width = "100%",   # constrain to page width, prevents overflow
  warning = FALSE,
  message = FALSE
)
```

## Installation

Install mongolstats, and mongolmaps for the maps, from GitHub:

```{r install}
# install.packages("pak")
# pak::pak("temuulene/mongolstats")
# pak::pak("temuulene/mongolmaps")
```

## Your First Analysis: Infant Mortality

This walk-through uses the infant mortality rate (IMR), a key indicator of population health and health system performance. It covers the full workflow: find a table, check its structure, fetch data, calculate rates, and map them.

### Step 1: Load Packages

```{r setup}
library(mongolstats)
library(mongolmaps)
library(dplyr)
library(ggplot2)

# Set language to English for readable output
nso_options(mongolstats.lang = "en")

# Set global ggplot2 theme with proper margins to prevent text cutoff
theme_set(
  theme_minimal(base_size = 11) +
    theme(
      plot.margin = margin(10, 10, 10, 10),
      plot.title = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 10, color = "grey40"),
      legend.text = element_text(size = 9),
      legend.title = element_text(size = 10)
    )
)
```

### Step 2: Find the Right Table

Search the catalogue by keyword:

```{r search}
mortality_tables <- nso_search("infant mortality")
mortality_tables |>
  distinct(tbl_id, tbl_eng_nm)
```

NSO publishes both **rates** (per 1,000 live births) and the **counts** behind them (infant deaths). We start with the monthly rate table, `DT_NSO_2100_015V1`, and later use the counts to compare aimags.

### Step 3: Explore Table Metadata

Before fetching data, check which dimensions the table has:

```{r metadata}
# View table structure
meta <- nso_table_meta("DT_NSO_2100_015V1")
meta

# Periods oldest first, with their dates. Month labels are "YYYY-MM"; the
# codes count back from the latest month ("0"), so select periods by label.
months <- nso_table_periods("DT_NSO_2100_015V1")
head(months)
range(months$label)
```

### Step 4: Fetch Data

Get the national rate for every month NSO publishes:

```{r fetch-data}
imr_national <- nso_data(
  tbl_id = "DT_NSO_2100_015V1",
  selections = list(
    Region = "0", # national total
    Month = months$label
  ),
  labels = "en" # add English label columns
)

# The Month column holds labels such as "2024-03", which keep their meaning
# when NSO publishes a new month
imr_national |>
  arrange(Month) |>
  head(10)
```

NSO rounds these monthly rates to whole numbers, which is coarse for a measure that sits around 10 to 20 per 1,000.

### Step 5: Visualize the Trend

```{r plot-trend, fig.alt="Line plot of Mongolia's monthly infant mortality rate since 2016, with a LOESS smoother showing the overall direction"}
imr_national |>
  mutate(date = nso_period_date(Month)) |> # "YYYY-MM" to Date
  ggplot(aes(x = date, y = value)) +
  geom_line(color = "#2c3e50", linewidth = 1, alpha = 0.3) + # dim raw data so trend stands out
  geom_point(color = "#e74c3c", size = 2, shape = 21, fill = "white", stroke = 1, alpha = 0.6) +
  # LOESS reveals the trend through month-to-month noise
  geom_smooth(
    method = "loess", formula = y ~ x, se = TRUE,
    color = "#3498db", fill = "#3498db", alpha = 0.2, linewidth = 1.5
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = "Infant Mortality Rate in Mongolia (Monthly)",
    subtitle = "Deaths under age 1 per 1,000 live births, national",
    x = NULL,
    y = "IMR (per 1,000)",
    caption = "Source: NSO Mongolia, DT_NSO_2100_015V1"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank() # vertical gridlines clutter time series
  )
```

## Regional Comparison

To compare aimags over a year, do not average the 12 monthly rates. A mean of monthly rates gives a month with few births the same weight as a busy month, and NSO's rounding makes small aimags' monthly rates jump between values. Instead, add up the infant deaths and the live births for the year, then divide:

$$\text{IMR} = \frac{\text{infant deaths in the year}}{\text{live births in the year}} \times 1{,}000$$

The counts come from two monthly tables: infant deaths (`DT_NSO_2100_014V1`) and live births (`DT_NSO_2100_018V5`).

```{r regional}
# Sum a monthly count table over one calendar year, for every region code
annual_counts <- function(tbl, year) {
  months <- nso_dim_values(tbl, "Month", labels = "en")
  nso_data(
    tbl_id = tbl,
    selections = list(
      Region = nso_dim_values(tbl, "Region")$code,
      Month = months$code[startsWith(months$label_en, year)]
    ),
    labels = "en"
  ) |>
    summarise(count = sum(value), n_months = n(), .by = c(Region, Region_en))
}

# Latest year with all 12 months published
birth_months <- nso_dim_values("DT_NSO_2100_018V5", "Month", labels = "en")$label_en
latest_year <- names(which(table(substr(birth_months, 1, 4)) == 12)) |> max()

infant_deaths <- annual_counts("DT_NSO_2100_014V1", latest_year)
live_births <- annual_counts("DT_NSO_2100_018V5", latest_year)

imr_regional <- infant_deaths |>
  select(Region, Region_en, deaths = count) |>
  inner_join(select(live_births, Region, births = count), by = join_by(Region)) |>
  mutate(
    Region_en = trimws(Region_en),
    level = case_when(
      Region == "0" ~ "National",
      Region %in% c("1", "2", "3", "4") ~ "Region",
      # Ulaanbaatar's figures sit under code "5"; code "511" is an empty duplicate
      Region == "5" ~ "Capital",
      Region == "511" ~ "Duplicate",
      .default = "Aimag"
    ),
    imr = deaths / births * 1000,
    # Exact Poisson 95% confidence interval for the rate
    lower = qchisq(0.025, 2 * deaths) / 2 / births * 1000,
    upper = qchisq(0.975, 2 * (deaths + 1)) / 2 / births * 1000
  )

national_imr <- imr_regional$imr[imr_regional$level == "National"]

imr_units <- imr_regional |>
  filter(level %in% c("Aimag", "Capital"))

# Highest rates
imr_units |>
  arrange(desc(imr)) |>
  select(Region_en, deaths, births, imr, lower, upper) |>
  head(10)
```

In `r if (run_vignette) latest_year else "the latest year"`, the national IMR was `r if (run_vignette) round(national_imr, 1) else "-"` per 1,000 live births. Many aimags record only a few dozen infant deaths a year, so their rates come with wide confidence intervals.

### Visualize Regional Differences

```{r regional-plot, fig.alt="Dot plot of infant mortality rates by aimag and Ulaanbaatar for the latest complete year, with 95% confidence intervals and a dashed line at the national rate", fig.height=7}
imr_units |>
  mutate(Region_en = forcats::fct_reorder(Region_en, imr)) |> # order by rate, not alphabet
  ggplot(aes(x = imr, y = Region_en)) +
  geom_vline(xintercept = national_imr, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(xmin = lower, xmax = upper), width = 0.3, color = "grey60", orientation = "y") +
  geom_point(aes(color = level), size = 2.5) +
  scale_color_manual(values = c(Aimag = "#2a78d6", Capital = "#eb6834"), name = NULL) +
  labs(
    title = paste0("Infant Mortality by Aimag, ", latest_year),
    subtitle = "Per 1,000 live births, with exact 95% CIs; dashed line = national rate",
    x = "Deaths per 1,000 live births",
    y = NULL,
    caption = "Source: NSO Mongolia, DT_NSO_2100_014V1 and DT_NSO_2100_018V5"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

Where an interval crosses the dashed line, that aimag's rate is consistent with the national rate. Infant deaths and births are counted where they are recorded, so referrals of high-risk births to hospitals in Ulaanbaatar can raise the capital's rate and lower the rate of the aimag the mother came from.

## Adding Geographic Context

[mongolmaps](https://github.com/temuulene/mongolmaps) provides boundaries that carry NSO codes, so the `Region` column joins without matching names.

```{r map-example, fig.alt="Choropleth map of infant mortality rates across Mongolia's aimags and Ulaanbaatar for the latest complete year"}
# mn_join() takes Ulaanbaatar's figures from region code "5" when the
# capital's own code "511" is empty, as it is in this table
imr_map <- imr_units |>
  mn_join(by = "Region", level = "aimag")

mn_map(
  imr_map,
  fill = imr,
  title = paste0("Infant Mortality Rate, ", latest_year)
) +
  labs(subtitle = "Infant deaths per 1,000 live births. Data: NSO Mongolia", fill = "IMR\n(per 1,000)")
```

The map shows point estimates only. Read it with the confidence intervals above: in small aimags, a few more or fewer infant deaths can move the colour a long way.

## Key Functions Summary

| Function | Purpose | Example |
|----------|---------|---------|
| `nso_search()` | Find tables by keyword | `nso_search("mortality")` |
| `nso_table_meta()` | Get table dimensions | `nso_table_meta("DT_NSO_...")` |
| `nso_dim_values()` | List dimension values | `nso_dim_values(tbl, "Region")` |
| `nso_table_periods()` | Check time coverage | `nso_table_periods(tbl)` |
| `nso_latest_periods()` | Latest periods, for selections | `list(Year = nso_latest_periods(tbl, n = 5))` |
| `nso_data()` | Fetch data | `nso_data(tbl, selections, labels)` |
| `nso_period_date()` | Period labels to dates | `nso_period_date(data$Month)` |
| `mongolmaps::mn_join()` | Join NSO data to boundaries by code | `mn_join(data, by = "Region", level = "aimag")` |
| `mongolmaps::mn_map()` | Quick map of joined data | `mn_map(joined, fill = value)` |

## Best Practices

1. **Use codes in scripts, labels for display**: Selections accept codes or labels, but codes such as region codes are more stable. `labels = "en"` adds readable label columns alongside the codes. Periods are the exception: NSO numbers them from the latest (`"0"`), so mongolstats always returns them as labels, and you should select them by label or with `nso_latest_periods()`.
2. **Check metadata first**: Use `nso_table_meta()` and `nso_dim_values()` to see dimensions and codes before fetching.
3. **Separate the levels**: Region columns mix the national total (`"0"`), regions (`"1"`–`"4"`), the capital (`"5"`), and aimags. Filter to one level before summing or ranking, or you will count deaths twice.
4. **Build rates from counts**: Add up numerators and denominators over the period, then divide. Do not average monthly rates.
5. **Show uncertainty**: Small populations give unstable rates. Report confidence intervals, or pool several years.
6. **Join by code**: Use `mongolmaps::mn_join()` on the `Region` code column rather than matching names.

## Common Workflows

### Time Series Analysis
1. Search for table → Check periods → Fetch months or years → Plot trend

### Regional Comparison
1. Fetch numerator and denominator counts → Sum over the period → Calculate rates and confidence intervals → Compare with the national rate

### Spatial Epidemiology
1. Calculate rates by aimag → `mn_join()` by `Region` code → `mn_map()`

## Next Steps

- **Discover More Data**: See the [Discovery Guide](discovery.html) for advanced search techniques
- **Create Maps**: Learn spatial analysis in the [Mapping Guide](mapping.html)
- **Reference**: Browse all functions in the [Reference](https://temuulene.github.io/mongolstats/reference/index.html)

## Quick Reference: Common Health Tables

```{r health-tables, echo=FALSE}
tibble::tribble(
  ~Indicator, ~Table_ID, ~Measure, ~Detail,
  "Infant mortality rate", "DT_NSO_2100_015V1", "Rate per 1,000 live births", "Aimag, month",
  "Infant deaths", "DT_NSO_2100_014V1", "Count", "Aimag, month",
  "Live births", "DT_NSO_2100_018V5", "Count", "Aimag, month",
  "Under-5 mortality rate", "DT_NSO_2100_030V2", "Rate per 1,000 live births", "Aimag, month",
  "Maternal deaths", "DT_NSO_2100_002V2", "Count", "Aimag, month",
  "Deaths by leading cause", "DT_NSO_2100_027V1", "Count", "Soum, year",
  "Cancer incidence", "DT_NSO_2100_012V1", "Rate per 10,000 population", "Cancer type, year",
  "Communicable disease cases (incl. tuberculosis)", "DT_NSO_2100_035V1", "Count", "Disease, month",
  "Communicable disease cases", "DT_NSO_2100_020V2", "Count", "Disease, soum, year",
  "Mid-year population", "DT_NSO_0300_002V1", "Count", "Aimag, year"
) |>
  knitr::kable()
```
