---
title: "Ulaanbaatar District-Level Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ulaanbaatar District-Level Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r not-cran-check, include = FALSE}
# Skip code evaluation on CRAN (external API calls will fail)
not_cran_check <- identical(Sys.getenv("NOT_CRAN"), "true") &&
  !nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
run_vignette <- isTRUE(getOption("pkgdown.in_progress")) || not_cran_check
knitr::opts_chunk$set(eval = run_vignette, purl = run_vignette)
# Inline `r` code runs even when chunks are skipped (eval = FALSE), so
# values computed in chunks go through this helper. `expr` is only
# evaluated when the chunks ran, and falls back if it cannot be computed.
inline_value <- function(expr, fallback = "n/a") {
  if (!run_vignette) {
    return(fallback)
  }
  tryCatch(expr, error = function(e) fallback)
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,        # standardized width that fits vignette text column
  fig.height = 5,       # default height, override in individual chunks as needed
  fig.dpi = 150,        # higher resolution for crisp rendering
  out.width = "100%",   # constrain to page width, prevents overflow
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(mongolstats)
library(mongolmaps)
library(dplyr)
library(ggplot2)
library(sf)
library(tidyr)

nso_options(mongolstats.lang = "en")

# Global theme with proper margins to prevent text cutoff
theme_set(
  theme_minimal(base_size = 11) +
    theme(
      plot.margin = margin(10, 10, 10, 10),
      plot.title = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 10, color = "grey40"),
      legend.text = element_text(size = 9),
      legend.title = element_text(size = 10)
    )
)
```

## Introduction

Ulaanbaatar, the capital of Mongolia, is home to about half of the country's population. This vignette looks at its nine districts (duureg) using NSO tables published "by soum and district": population and growth, school capacity, the air quality monitoring network, and the limits of district-level health statistics.

## Districts and Their Codes

[mongolmaps](https://github.com/temuulene/mongolmaps) provides the district boundaries with their NSO codes and areas, so NSO tables join without matching names.

```{r helpers}
ub_districts <- mn_ub_districts()
ub_districts |>
  st_drop_geometry() |>
  select(name, nso_code, area_km2)

# Fetch an NSO table for the nine districts only
get_ub_data <- function(tbl_id, selections = list()) {
  nso_data(
    tbl_id,
    selections = c(list(Region = ub_districts$nso_code), selections),
    labels = "en"
  ) |>
    mutate(District = trimws(Region_en))
}

# Latest year a table covers
latest_year <- function(tbl_id) {
  max(nso_dim_values(tbl_id, "Year", labels = "en")$label_en)
}
```

## 1. Demographic Profile

### Population Density

The population table `DT_NSO_0300_002V4` gives the mid-year resident population down to khoroo level. (This table id names two tables in the catalogue; mongolstats warns and uses the first match, which is the khoroo-level table and also includes the districts.)

```{r demog-density, fig.alt="Map of population density by Ulaanbaatar district on a log scale, with each district labelled"}
pop_tbl <- "DT_NSO_0300_002V4"
pop_year <- latest_year(pop_tbl)

pop_data <- get_ub_data(pop_tbl, list(Year = pop_year))

ub_pop_map <- pop_data |>
  mn_join(by = "Region", level = "soum", within = "Ulaanbaatar") |>
  mutate(density = value / area_km2) # residents per km2

mn_map(
  ub_pop_map,
  fill = density, trans = "log10", label = TRUE,
  title = paste0("Population Density by District, ", pop_year)
) +
  labs(
    subtitle = "Residents per km² (log scale). Data: NSO Mongolia, DT_NSO_0300_002V4",
    fill = "Residents\nper km²"
  )
```

```{r density-facts, include = FALSE}
density_rank <- if (run_vignette) {
  ub_pop_map |>
    st_drop_geometry() |>
    arrange(desc(density))
}
```

**`r inline_value(density_rank$District[1])`** is the densest district, with about `r inline_value(scales::label_comma()(round(density_rank$density[1], -2)))` residents per km². **`r inline_value(density_rank$District[9])`** is the least dense, at about `r inline_value(round(density_rank$density[9]))` per km². Districts such as Bayanzurkh and Songinokhairkhan include large areas of mountain and pasture, so a district average hides very dense neighbourhoods. Khoroo-level data show more detail:

```{r khoroo-density, fig.alt="Map of population density by khoroo in Ulaanbaatar's six central districts on a log scale", fig.height=5.5}
khoroo_codes <- nso_dim_values(pop_tbl, "Region")$code |>
  stringr::str_subset("^511\\d{4}$") # seven-digit codes are khoroos

khoroo_pop <- nso_data(
  pop_tbl,
  selections = list(Region = khoroo_codes, Year = pop_year),
  labels = "en"
) |>
  mn_join(by = "Region", level = "bag", within = "Ulaanbaatar")

central <- c("Bayangol", "Bayanzurkh", "Chingeltei", "Khan-Uul", "Songinokhairkhan", "Sukhbaatar")
mn_map(
  filter(khoroo_pop, parent_pcode %in% ub_districts$pcode[ub_districts$name %in% central]),
  fill = value / area_km2, trans = "log10",
  title = paste0("Population Density by Khoroo, Central Ulaanbaatar, ", pop_year)
) +
  labs(subtitle = "Residents per km² (log scale), six central districts", fill = "Residents\nper km²")
```

The khoroo boundaries in mongolmaps come from an open project that does not state their source, so treat them as unofficial.

### Population Growth Trends

Rapid growth can strain local health and school services.

```{r demog-trend, fig.alt="Line chart of resident population by Ulaanbaatar district over the last 10 years"}
trend_years <- as.character(as.integer(pop_year) - 9:0)
pop_trend <- get_ub_data(pop_tbl, list(Year = trend_years)) |>
  mutate(Year = as.integer(Year_en))

pop_trend |>
  ggplot(aes(x = Year, y = value, color = District)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = as.integer(trend_years)) +
  labs(
    title = paste0("Resident Population by District, ", min(trend_years), "–", max(trend_years)),
    subtitle = "Mid-year resident population",
    y = "Population", x = NULL, color = "District",
    caption = "Source: NSO Mongolia, DT_NSO_0300_002V4"
  ) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r growth-facts, include = FALSE}
growth <- if (run_vignette) {
  pop_trend |>
    filter(Year %in% range(Year)) |>
    summarise(
      change = (value[Year == max(Year)] / value[Year == min(Year)] - 1) * 100,
      .by = District
    ) |>
    arrange(desc(change))
}
fmt_change <- function(x) sprintf("%+.0f%%", x)
```

Over these 10 years, **`r inline_value(growth$District[1])`** grew fastest (`r inline_value(fmt_change(growth$change[1]))`), followed by **`r inline_value(growth$District[2])`** (`r inline_value(fmt_change(growth$change[2]))`). **`r inline_value(growth$District[9])`** changed the least (`r inline_value(fmt_change(growth$change[9]))`). Growth this uneven means services planned on older population figures may now be under strain in the fastest-growing districts.

## 2. School Capacity

NSO publishes few socioeconomic indicators by district. The number of students per general education school gives a rough measure of pressure on public services.

```{r socio-edu, fig.alt="Bar chart of students per general education school by Ulaanbaatar district"}
edu_year <- latest_year("DT_NSO_2002_057V2")
schools <- get_ub_data("DT_NSO_2001_002V2", list(Year = edu_year)) |>
  select(District, Schools = value)
students <- get_ub_data("DT_NSO_2002_057V2", list(Year = edu_year)) |>
  select(District, Students = value)

edu_load <- inner_join(schools, students, by = join_by(District)) |>
  mutate(Students_Per_School = Students / Schools)

ggplot(edu_load, aes(x = Students_Per_School, y = reorder(District, Students_Per_School))) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_text(aes(label = round(Students_Per_School)), hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::label_comma(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = paste0("Students per General Education School, ", edu_year),
    subtitle = "Full-time students divided by number of schools",
    x = "Students per school", y = NULL,
    caption = "Source: NSO Mongolia, DT_NSO_2001_002V2 and DT_NSO_2002_057V2"
  ) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
```

Schools differ in size and shift system, so a high ratio points to crowding but does not measure it directly.

## 3. Health Statistics by District: A Caution

Rates need a numerator and a denominator for the **same population**. NSO's district tables count residents by where they live, but births and deaths can be counted where they happen, which is often a hospital in another district. A quick check is to divide each district's births and deaths by its residents:

```{r health-check, fig.alt="Two-panel bar chart of live births and deaths per 1,000 residents by Ulaanbaatar district, showing implausible differences between districts"}
vital_year <- latest_year("DT_NSO_2100_027V1")

births <- get_ub_data(
  "DT_NSO_2100_018V6",
  list(Indicator = "0", Year = vital_year) # Indicator "0" = both sexes
) |>
  select(District, count = value) |>
  mutate(measure = "Live births per 1,000 residents")

deaths <- get_ub_data(
  "DT_NSO_2100_027V1",
  list(Indicator = "0", Year = vital_year) # Indicator "0" = all deaths
) |>
  select(District, count = value) |>
  mutate(measure = "Deaths per 1,000 residents")

residents <- get_ub_data(pop_tbl, list(Year = vital_year)) |>
  select(District, residents = value)

vital_rates <- bind_rows(births, deaths) |>
  inner_join(residents, by = join_by(District)) |>
  mutate(rate = count / residents * 1000)

vital_rates |>
  ggplot(aes(x = rate, y = reorder(District, residents))) +
  geom_col(fill = "#8e44ad", width = 0.7) +
  geom_text(aes(label = round(rate, 1)), hjust = -0.1, size = 3) +
  facet_wrap(~measure, scales = "free_x") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = paste0("Recorded Births and Deaths per 1,000 Residents, ", vital_year),
    subtitle = "Districts ordered by number of residents",
    x = NULL, y = NULL,
    caption = "Source: NSO Mongolia, DT_NSO_2100_018V6, DT_NSO_2100_027V1 and DT_NSO_0300_002V4"
  ) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
```

```{r health-facts, include = FALSE}
birth_rates <- if (run_vignette) filter(vital_rates, startsWith(measure, "Live births")) |> arrange(rate)
death_rates <- if (run_vignette) filter(vital_rates, startsWith(measure, "Deaths")) |> arrange(rate)
```

These numbers cannot describe the residents. **`r inline_value(birth_rates$District[1])`**, with `r inline_value(scales::label_comma()(birth_rates$residents[1]))` residents, recorded `r inline_value(birth_rates$count[1])` live births in `r inline_value(vital_year)`, while **`r inline_value(birth_rates$District[9])`** recorded `r inline_value(round(birth_rates$rate[9]))` per 1,000 residents. Recorded deaths range from `r inline_value(round(death_rates$rate[1], 1))` per 1,000 residents in `r inline_value(death_rates$District[1])` to `r inline_value(round(death_rates$rate[9], 1))` in `r inline_value(death_rates$District[9])`. The pattern fits counting by **place of occurrence**: births and deaths are recorded in the districts that have maternity and general hospitals, not in the districts where the mother or the person who died lived.

So district-level rates built from these tables, such as infant mortality or deaths from respiratory disease per 10,000 residents, mix up where people live with where hospitals are. Correlating them with district density or other indicators would produce misleading results. For district comparisons, use data coded by **place of residence**, which may need a request to the Health Development Center or NSO. City-wide rates for Ulaanbaatar as a whole are less affected, though referrals of patients from the aimags to the capital's hospitals still raise its counts.

## 4. Air Quality Monitoring Network

Station-level air quality data (see the [Environmental Analysis](environmental-surveillance.html) vignette) can only be linked to people if we know where the stations are. NSO does not publish station coordinates, so the table below lists approximate locations for 18 Ulaanbaatar stations, compiled from public station listings, including the World Air Quality Index project (api.waqi.info). Check them before using them for exposure assignment.

```{r aq-station-map}
ub_aq_stations <- tribble(
  ~Station, ~Latitude, ~Longitude,
  "Misheel-Expo center", 47.89428, 106.8825,
  "West crossroad", 47.91576, 106.8940,
  "1st micro district", 47.91798, 106.8481,
  "13th micro district", 47.91761, 106.9374,
  "32nd Toirog", 47.939333, 106.914333,
  "Ofitseruudiin ordon", 47.916611, 106.972556,
  "Kharkhorin market", 47.910583, 106.837111,
  "Urgakh naran micro district", 47.873778, 107.115028,
  "Dambdarjaa", 47.988250, 106.951167,
  "Khailaast", 47.963528, 106.893889,
  "Tolgoit", 47.920278, 106.799250,
  "Zuragt", 47.92973, 106.8886,
  "Amgalan", 47.906250, 107.016556,
  "Nisekh", 47.86394, 106.7791,
  "Tavan buudal", 47.954500, 106.914833,
  "Bayankhoshuu", 47.945528, 106.822917,
  "Sharkhad", 47.93375, 107.0103,
  "100 ail", 47.93291, 106.9214
)

# Assign each station to the district polygon it falls in
ub_stations_sf <- ub_aq_stations |>
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) |>
  st_join(select(ub_districts, District = name))

# Station count for every district, including those with none
station_count <- ub_districts |>
  st_drop_geometry() |>
  select(District = name) |>
  left_join(count(st_drop_geometry(ub_stations_sf), District, name = "Stations"), by = join_by(District)) |>
  mutate(Stations = coalesce(Stations, 0L)) |>
  arrange(desc(Stations))

knitr::kable(station_count, caption = "Air quality monitoring stations by district")
```

Stations are assigned by their coordinates, using mongolmaps' simplified district boundaries. A station within a few hundred metres of a district border may fall on the other side.

### Spatial Distribution

```{r aq-network-map, fig.height=6, fig.alt="Map of Ulaanbaatar's central districts with air quality monitoring stations shown as red points"}
# Zoom to the stations, with a 3 km margin, in UTM zone 48N (metres)
utm48n <- 32648
station_box <- ub_stations_sf |>
  st_transform(utm48n) |>
  st_buffer(3000) |>
  st_bbox()

mn_map(filter(ub_districts, name %in% central), label = TRUE, crs = utm48n,
       title = "Air Quality Monitoring Stations in Central Ulaanbaatar") +
  geom_sf(data = ub_stations_sf, color = "#e34948", size = 2.5, alpha = 0.9) +
  ggrepel::geom_text_repel(
    data = ub_stations_sf, aes(label = Station, geometry = geometry), stat = "sf_coordinates",
    size = 2.4, color = "grey20", min.segment.length = 0, seed = 1
  ) +
  coord_sf(
    crs = utm48n, datum = NA,
    xlim = station_box[c("xmin", "xmax")], ylim = station_box[c("ymin", "ymax")]
  ) +
  labs(subtitle = "Red points = monitoring stations (approximate locations)")
```

### Coverage

```{r station-coverage, fig.alt="Bar chart of residents per monitoring station for each Ulaanbaatar district that has at least one station"}
station_coverage <- station_count |>
  left_join(select(pop_data, District, Population = value), by = join_by(District)) |>
  mutate(Pop_per_Station = if_else(Stations > 0, Population / Stations, NA_real_))

station_coverage |>
  filter(Stations > 0) |>
  ggplot(aes(x = Pop_per_Station / 1000, y = reorder(District, Pop_per_Station))) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_text(aes(label = paste0(round(Pop_per_Station / 1000), "k (", Stations, ")")), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Residents per Monitoring Station",
    subtitle = "Thousands of residents per station (number of stations in brackets)",
    x = "Residents per station (thousands)", y = NULL,
    caption = "Districts with no station are not shown"
  ) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
```

```{r coverage-facts, include = FALSE}
no_station <- if (run_vignette) filter(station_coverage, Stations == 0)
```

`r inline_value(nrow(no_station))` districts, `r inline_value(knitr::combine_words(no_station$District))`, have no station in this network. About `r inline_value(scales::label_comma()(round(sum(no_station$Population), -3)))` residents (`r inline_value(round(sum(no_station$Population) / sum(station_coverage$Population) * 100))`% of the city) live there. Residents per station is a crude measure: pollution varies with distance from sources and with terrain, so the location of stations matters as much as their number. The ger areas on the city's northern and eastern edges, where most household coal and wood burning happens, are where denser monitoring would add the most to exposure estimates.

## Conclusion

Ulaanbaatar's districts differ widely in density, growth and school capacity, and NSO's district tables plus mongolmaps' boundaries make these differences easy to map. Health outcomes are harder: published district births and deaths follow hospital locations, so district-level rates need data coded by place of residence before they can guide local public health action.
