---
title: "Fuzzy name joins with mongolstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fuzzy name joins with mongolstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
eval_site <- TRUE
if (requireNamespace("pkgdown", quietly = TRUE)) eval_site <- pkgdown::in_pkgdown()
if (eval_site && requireNamespace("curl", quietly = TRUE)) eval_site <- curl::has_internet()
knitr::opts_chunk$set(eval = eval_site, message = FALSE, warning = FALSE, error = TRUE)
if (eval_site) {
  options(mongolstats.retry_tries = 5L, mongolstats.timeout = 60)
  try(mongolstats::nso_cache_enable(), silent = TRUE)
}
# Always install local package before knitting to ensure latest API
if (requireNamespace("remotes", quietly = TRUE)) {
  try(remotes::install_local(".", upgrade = "never", force = TRUE, dependencies = FALSE), silent = TRUE)
}
```

```{r}
library(mongolstats)
library(dplyr)
library(sf)
```

This vignette shows how to join tabular data to ADM1 boundaries using fuzzy matching.

```{r}
adm1 <- mn_boundaries("ADM1") %>% mn_boundaries_normalize()

# Create a small example with slightly altered names
sample_names <- head(adm1$shapeName, 6)
altered <- sub("a", "", sample_names) # remove one 'a' to simulate typos
dat <- tibble::tibble(name = altered, value = seq_along(altered))

# Fuzzy join (max distance 2)
joined <- mn_fuzzy_join_by_name(dat, name_col = "name", level = "ADM1", boundaries = adm1, max_distance = 2)
joined %>% select(shapeName, name, value) %>% head()
```

If you have official region names in your dataset, prefer `mn_join_by_name()` for exact matching.

## Guidance for fuzzy matching

- Use `max_distance` conservatively (e.g., 1–2) to avoid incorrect matches.
- You can supply a different string distance `method` if `stringdist` is installed; Jaro-Winkler and Levenshtein are common choices.
- Always review near ties and ambiguous matches manually; consider exporting candidate pairs for QA if the dataset is critical.
